---
title: "Creating ADCE"
output: 
   rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating ADCE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r setup, include = FALSE}
# Other variables
admiral_homepage <- "https://pharmaverse.github.io/admiral"
library(admiraldev)
```

# Introduction

This article describes creating an `ADCE` ADaM for the analysis of Vaccine Reactogenicity Data collected in SDTM `CE` domain. The current  
presented example is tested using `CE`  SDTM domains and `ADSL` ADaM domain. However, other domains could be used if needed (eg temperature data collected in `VS`).

**Note:** *All examples assume CDISC SDTM and/or ADaM format as input unless 
otherwise specified.*

# Programming Flow
* [Read in Data](#readdata)
* [Restrict CE to Reactogenicity records](#cefilter)
* [Creation of Periods Reference dataset](#adperiods)
* [Derivation of Analysis Dates](#adates)
* [Join with the Periods Reference dataset](#mergeperiods)
* [Creation of analysis version for GRADING Variable](#agrade)
* [Creation of Analysis Sequence Number](#aseq)
* [Final Step : get all the remaining variables from ADSL](#jadsl)

## Read in Data {#readdata}
Assumption: The CE domain has already been merged with the SUPPCE dataset. If this is not yet the case, join SUPPCE onto parent CE domain using metatools::combine_supp(`CE`, `SUPPCE`).

```{r, eval=TRUE, message=FALSE, warning=FALSE}
library(admiraldev)
library(admiral)
library(dplyr)
library(lubridate)
library(admiralvaccine)

data("vx_ce")
data("vx_adsl")


adsl <- vx_adsl
ce <- vx_ce

ce <- convert_blanks_to_na(ce)
adsl <- convert_blanks_to_na(adsl)
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adsl,
  display_vars = exprs(USUBJID, TRTSDT, TRTEDT, TRT01A, AP01SDT, AP01EDT, AP02SDT, AP02EDT)
)
```

## Restrict CE domain to the relevant records for Reactogenicity Analysis {#cefilter}
```{r eval=TRUE}
# Get CE records
adce <- ce %>%
  filter(CECAT == "REACTOGENICITY")
```

## Create Reference Dataset for Periods {#adperiods}
```{r eval=TRUE}
# Create period dataset - for joining period information onto ce records
# Need to remove datetime variables as otherwise causes duplicate issues
adsl2 <- adsl %>%
  select(-c(starts_with("AP") & ends_with("DTM")))

adperiods <- create_period_dataset(
  adsl2,
  new_vars = exprs(APERSDT = APxxSDT, APEREDT = APxxEDT)
)
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adperiods,
  display_vars = exprs(USUBJID, APERIOD, APERSDT, APEREDT)
)
```

## Select predecessor variables from ADSL, and derive analysis dates (ASTDT, ASTDY, AENDT, AENDY) {#adates}

At this step, it may be useful to join ADSL to your CE domain. Only the ADSL variables used for derivations are selected at this step. The rest of the relevant ADSL variables would be added later.

```{r eval=TRUE}
# Get list of ADSL vars required for derivations
adsl_vars <- exprs(TRTSDT, TRTEDT)
# Derive analysis dates/days
adce <- adce %>%
  # join adsl to ce
  derive_vars_merged(
    dataset_add = adsl,
    new_vars = adsl_vars,
    by = exprs(STUDYID, USUBJID)
  ) %>%
  ## Derive analysis start time ----
  ## Proposed imputations depending on situation: no needed -> highest imputation = “n”, 
  ## some missing dates -> highest imputation = “D”
  derive_vars_dt(
    dtc = CESTDTC,
    new_vars_prefix = "AST",
    highest_imputation = "n"
  ) %>%
  ## Derive analysis end time ----
  derive_vars_dt(
    dtc = CEENDTC,
    new_vars_prefix = "AEN",
    highest_imputation = "n"
  ) %>%
  ## Derive analysis start relative day and  analysis end relative day ----
  derive_vars_dy(
    reference_date = TRTSDT,
    source_vars = exprs(ASTDT, AENDT)
  )
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adce,
  display_vars = exprs(USUBJID, TRTSDT, CESTDTC, CEENDTC, ASTDT, AENDT, ASTDY, AENDY)
)
```

## Join with the periods reference dataset and derive Relative Day in Period {#mergeperiods}
Also add analysis version of CEREL(AREL).
```{r eval=TRUE}
adce <-
  derive_vars_joined(
    adce,
    dataset_add = adperiods,
    by_vars = exprs(STUDYID, USUBJID),
    filter_join = ASTDT >= APERSDT & ASTDT <= APEREDT
  ) %>%
  mutate(
    APERSTDY = as.integer(ASTDT - APERSDT) + 1,
    AREL = CEREL
  )
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adce,
  display_vars = exprs(USUBJID, TRTSDT, ASTDT, AENDT, ASTDY, AENDY, APERIOD, APERSDT, APERSTDY)
)
```

##  Creation of analysis version for GRADING Variable (either TOXGR or SEV) {#agrade}
Depending on which variable is collected for the Grading (TOXGR or SEV) in CE domain, derive the associated analysis version.
In current example, SEV is collected, so the code is using this as an example.
In addition, derivation of Extreme Flags: in current example: flag the first occurence of the most severe grade within a Period (AOCC01FL).

```{r eval=TRUE}
adce <- adce %>%
  ## depending on collection of TOXGR or SEV in CE domain ----
  ## Analysis variant of ASEV and ASEVN ----
  mutate(
    ASEV = CESEV,
    ASEVN = as.integer(factor(ASEV,
      levels = c("MILD", "MODERATE", "SEVERE", "DEATH THREATENING")
    ))
  ) %>%
  ## Derive occurrence flags: first occurrence of most severe solicited AE ----
  ## - Company specific ----
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(USUBJID, APERIOD),
      order = exprs(desc(ASEVN), ASTDY, CEDECOD),
      new_var = AOCC01FL,
      mode = "first"
    ),
    filter = !is.na(APERIOD) & !is.na(ASEV)
  )
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adce,
  display_vars = exprs(USUBJID, TRTSDT, ASTDT, APERIOD, APERSDT, 
                       APERSTDY, CEDECOD, ASEVN, AOCC01FL, CESEQ)
)
```

## Creation of Analysis Sequence Number {#aseq}
```{r eval=TRUE}
adce <- adce %>%
  ## Derive ASEQ ----
  derive_var_obs_number(
    new_var = ASEQ,
    by_vars = exprs(STUDYID, USUBJID),
    order = exprs(CEDECOD, CELAT, CETPTREF, APERIOD),
    check_type = "error"
  ) %>%
  ## Derive analysis duration (value and unit) ----
  derive_vars_duration(
    new_var = ADURN,
    new_var_unit = ADURU,
    start_date = ASTDT,
    end_date = AENDT,
    in_unit = "days",
    out_unit = "days",
    add_one = TRUE,
    trunc_out = FALSE
  )
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adce,
  display_vars = exprs(USUBJID, TRTSDT, ASTDT, APERIOD, APERSDT, 
                       APERSTDY, CEDECOD, ASEVN, AOCC01FL, CESEQ, ASEQ)
)
```
## Final Step : get all the remaining variables from ADSL {#jadsl_list}
## Get list of ADSL vars as per trial specific which needs to be adjusted when using the template
```{r eval=TRUE}
# Get list of ADSL vars, list is trial specific and needs to be adjusted when using the template
adsl_list <- adsl %>%
  select(STUDYID, USUBJID, TRT01A, TRT01P, AGE, AGEU, SEX, RACE, COUNTRY, ETHNIC, SITEID, SUBJID)


# Join ADSL_list with CE
adce <- adce %>%
  derive_vars_merged(
    dataset_add = adsl_list,
    by_vars = exprs(STUDYID, USUBJID)
  )
```
